Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-12 intro.tex
compressing stuff up through (including) the strong virtual pointsto def

At its core, the virtual memory functionality of modern CPUs is about \emph{location virtualization}: the memory locations
(addresses) seen by most code are not, in fact, the exact location in physical memory where data reside. Instead these 
are \emph{virtual} addresses, which are mapped to actual physical resources by the cooperation of the hardware and OS. 
This is what enables separation of process memory resources:
the OS manipulates hardware functionality to ensure that any attempt by a process to access memory not explicitly granted 
to it by the kernel will fail. But this is complicated by the fact that 
%the OS and hardware can also enable, shared (overlapping) access to physical memory regions; the fact that the kernel data structures themselves are accessed via virtual memory addresses; 
%and the fact that 
control over these mappings of virtual to physical addresses is itself mediated by \emph{in-memory data structures}, 
which the kernel still accesses via virtual address, leading to indirect cycles.

Further complicating matters is that addresses themselves bear no information about which address space they originate 
from. For user processes this is of little concern, as these have access to only their own address space. But the kernel has
(or can grant itself) access to all address spaces. Mixing up addresses from different address spaces leads to severe bugs.
More concerning, keeping track of which \emph{assertions} hold in different address spaces during kernel verification is 
difficult: some assertions should hold across all address spaces, while others hold in only one, and others may hold in 
multiple but still not all.

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid
modal logic, which can label assertions true under \emph{other, named} circumstances (i.e., in another address space) with a 
modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a 
\textit{convenient} and \textit{powerful} way to \emph{modularly}
\begin{itemize}
\item isolate assertions specific to a particular address space,
\item explicitly state when an assertion is true across address spaces,
\item manipulate address spaces from within other address spaces, and
\item reason about change in address spaces.
\end{itemize}
These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, 
which were only able to work with a single address space (the current address space on the CPU) because they were unable
to speak directly \emph{within the logic} about other address spaces, in addition to handling
the \emph{non-local} effects of page table updates whether within the current address space or across address spaces.
%\begin{figure}
%   \includegraphics[width=0.5\columnwidth]{architecture.png}
%  \caption{Layers of Implementation}
%  \todo[inline]{Colin says: I don't think this figure helps readers; it is too abstract}
%  \label{fig:architecture}
%\end{figure}

\begin{itemize}
\item We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
      implemented as an embedded separation logic within the Iris~\cite{jung2018iris} separation logic.
      The result is a separation logic that lifts a number of major semantic restrictions present in the few
      prior logics tackling virtual address translation.
\item We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
\item We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, 
      including software page table walks, mapping  pages, and the assembly code for switching address spaces. 
      The latter in particular goes beyond the capabilities of prior work, and reveals unexpected
	subtleties.
\end{itemize}

The logic we develop here covers core reasoning principles for reasoning about memory configurations and code
reliant upon or manipulating those memory configurations in the presence of in-memory page tables, the primary
memory protection mechanism across Intel/AMD's x86-64 processors, ARM's application class processors including
aarch64 CPUs, POWER, RISC-V, and other architectures.
% One aspect of virtual memory management we do not address in this paper is updates to the translation lookaside
% buffer (TLB) present in most CPUs. Like CertiKOS~\cite{?}, seL4~\cite{sel4tocs}, and most other OS verification
% efforts we currently trust insertion of TLB flush operations, whose locations are critical but also few and
% well-known (after unmapping memory, or changing address spaces). Tackling correctness of TLB invalidation
% rigorously is an additional challenging project unto itself~\cite{...} with similarities to but
% significant divergences from reasoning about weak memory models, which future work should combine
% with the present work.

>>>>

At its core, the virtual memory functionality of modern CPUs is about \emph{location virtualization}: the memory locations
(addresses) seen by most code are not, in fact, the exact location in physical memory where data reside. Instead these 
are \emph{virtual} addresses, which are mapped to actual physical resources by the cooperation of the hardware and OS. 
This is what enables separation of process memory resources:
the OS manipulates hardware functionality to ensure that any attempt by a process to access memory not explicitly granted 
to it by the kernel will fail. But this is complicated by the fact that 
%the OS and hardware can also enable, shared (overlapping) access to physical memory regions; the fact that the kernel data structures themselves are accessed via virtual memory addresses; 
%and the fact that 
control over these mappings of virtual to physical addresses is itself mediated by \emph{in-memory data structures}, 
which the kernel still accesses via virtual address, leading to indirect cycles.
\looseness=-1

Further complicating matters, addresses themselves bear no information about which address space they originate 
from. 
%For user processes this is of little concern, as these have access to only their own address space. 
This is not a problem for usermode code, which only accesses its own address space.
But the kernel has
(or can grant itself) access to all address spaces. Mixing up addresses from different address spaces leads to severe bugs.
Proofs of kernel correctness must also track which \emph{assertions} hold in different address spaces:
some assertions should hold across all address spaces, while others hold in only one, and others may hold in 
multiple but still not all.

This kind of context-dependent assertion, where a fact may be true in one address space but not others, has a modal flavor. 
We propose tackling the verification of virtual memory subsystems (and kernels more broadly) by adapting ideas from hybrid
modal logic, which can label assertions true under \emph{other, named} circumstances (i.e., in another address space) with a 
modality indexed by a name for that space (in our case, the root of the page tables for an address space). This offers a 
\textit{convenient} and \textit{powerful} way to \emph{modularly}
isolate assertions specific to a particular address space,
explicitly state when an assertion is true across address spaces,
manipulate address spaces from within other address spaces, and
reason about change in address spaces.
These advantages make this approach to reasoning about virtual memory more flexible than prior program logic techniques~\cite{kolanski08vstte,kolanski09tphols}, 
which were only able to work with a single address space (the current address space on the CPU) because they were unable
to speak directly \emph{within the logic} about other address spaces, in addition to handling
the \emph{non-local} effects of page table updates whether within the current address space or across address spaces.

\paragraph{Contributions}
We develop these ideas in the form of a logic for working with virtual-address-space-relative assertions,
implemented as an embedded separation logic within Iris~\cite{jung2018iris}.
The result is a separation logic that lifts a number of major semantic restrictions present in the few
prior logics tackling virtual address translation.
The logic we develop covers core reasoning principles for reasoning about memory configurations and code
reliant upon or manipulating those memory configurations in the presence of in-memory page tables, the primary
memory protection mechanism across Intel/AMD's x86-64 processors, ARM's application class processors including
AArch64 CPUs, POWER, RISC-V, and other architectures.
We prove the soundness of our \textsf{vProp} logic with respect to a RISC-like fragment of \textsf{AMD64} instructions.
We verify simplified versions of several critical virtual-memory-related pieces of OS functionality, 
including mapping  pages, switching address spaces, and converting physical addresses to virtual addresses for
software page table walks (as when mapping pages). These examples demonstrate the suitability and flexibility of
a modal treatment of address spaces: each has a concise specification, each example either goes beyond the technical
capabilities of prior logics, or revisits an example from prior work with more details (treating parts of the kernel
invariant never addressed by prior work) and fewer assumptions (e.g., verifying virtual-to-physical conversion rather
than assuming it has already occurred).

% One aspect of virtual memory management we do not address in this paper is updates to the translation lookaside
% buffer (TLB) present in most CPUs. Like CertiKOS~\cite{?}, seL4~\cite{sel4tocs}, and most other OS verification
% efforts we currently trust insertion of TLB flush operations, whose locations are critical but also few and
% well-known (after unmapping memory, or changing address spaces). Tackling correctness of TLB invalidation
% rigorously is an additional challenging project unto itself~\cite{...} with similarities to but
% significant divergences from reasoning about weak memory models, which future work should combine
% with the present work.

### 2024-11-12 programlogic.tex
compressing stuff up through (including) the strong virtual pointsto def

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another variety of pointwise lifting that already exists in the main \iris
  distribution~\cite{dang2019rustbelt,dang2022compass}. This is essentially true, but the standard formalization does not work for index types which
  are indexed in the sense of \coq's metatheory, as our \lstinline|word n| types are.
}
\begin{lstlisting}[language=Coq]
  Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$.
\end{lstlisting}
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

\subsection{Base Points-To Assertions}
\label{sec:pointsto}
We build up our context-dependent assertions from two basic types of assertions that do not care about address spaces.

@@

\paragraph{Physical memory  points-to} The soundness proofs for our logic's rules largely center around 
proving that page-table-walk accesses as in Figure \ref{fig:pagetables} succeed, which requires assertions
dealing with physical memory locations.
We have two notions of physical points-to facts. The primitive notion closest to our machine model is captured by an assertion
$ \textsf{pfn} \ \sim \ \textsf{pageoff} \mapsto_{\textsf{a}} \; \{\textsf{q}\} \; \textsf{v} $, where \textsf{pfn} (a $\mathcal{W}_{52}$ \emph{page frame number}) essentially selects a 4KB page of physical memory,
and \textsf{pageoff} (a $\mathcal{W}_{12}$) is an offset within that page.
% could be an 52-bits masked address to level 4 table 
% ($\textsf{w1 } =( \textsf{ l4M52 maddr cr3val}) $),
%and, expectedly \textsf{w2} is an address computed by page-offset computation (e.g. $\textsf{l4off maddr cr3val}$). 
From this we can derive a more concise physical points-to when the split is unimportant:
% Giving a raw 64-bits memory pointsto assertion becomes
\[\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} \stackrel{\triangle}{=} (\textsf{drop 12}~w) \ \sim \ (\textsf{bottom 12}~w)\mapsto_{\textsf{a}} \; \{\textsf{q}\} \textsf{ v} \]
 %I put a newline in between the following two Definitions as the second Definition seems not proper without the newline
\begin{figure*}
  \begin{lstlisting}[language=Coq]
 Definition $\vaddr\mapsto_{\textsf{t}}\{\textsf{q}\}\; \vpage$ : vProp $\Sigma$ := 
 $\exists_{\textsf{l4e l3e l2e l1e}} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo($\vaddr$ l4e l3e l2e l1e paddr) $\ast$ paddr $\mapsto_{p}\{\mathsf{q}\} \vpage$.

 Definition L4_L1_PointsTo (maddr l4e l3e l2e l1e paddr :word 64) : vProp $\Sigma$ := $\lambda$ cr3val.
  $\ulcorner$ entry_present l4e $\land$ entry_present l3e $\land$ entry_present l2e $\land$ entry_present l1e$\urcorner$ $\ast$
  (l4M52 maddr cr3val) $\sim$ (l4off maddr cr3val) $\mapsto_{a}$ {q1}  l4e  $\ast$
  (l3M52 maddr l4e) $\sim$ (l3off maddr l4e)  $\mapsto_{a}$ {q2}  l3e $\ast$ 
  (l2M52 maddr l3e) $\sim$ (l2off maddr l3e) $\mapsto_{a}$ {q3}  l2e $\ast$
  (l1M52 maddr l2e) $\sim$ (l1off maddr l2e) $\mapsto_{a}$ {q4}  l1e $\ast$ $\ulcorner$ addr_L1(va,l1e) = paddr $\urcorner$.
\end{lstlisting}
\vspace{-1em}
\caption{A Strong Virtual Points-to Relation}
  \label{fig:strongvirtualpointsto}
\end{figure*}

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to that depends on the current address space is to
require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion contains
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk.
Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
to index each level, finding the existentially quantified entries in each level, 
ending with the L1 entry, whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.

\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry.\footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry.
}
For example, 
  the first level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the 
  physical address of the start of the L4 table,
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then at that physical location is the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized.\footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also a more explicit \textsf{vProp}, explicitly binding a value for \lstinline|cr3| and using it
to start the translation process.

>>>>

% The idea that the truth of an assertion is relative to an address space has far-reaching consequences.
To support making assertions depend on a choice of address space, we work entirely in a pointwise lifting of Iris's base BI logic,
essentially working with separation logic assertions indexed by a choice of page table root as a $\mathcal{W}_{64}$, which we call $\textsf{vProp }\Sigma$:\footnote{
  \iris experts may notice our \lstinline|-b>| resembles another variety of pointwise lifting that already exists in the main \iris
  distribution~\cite{dang2019rustbelt,dang2022compass}. This is essentially true, but the standard formalization does not work for index types which
  are indexed in the sense of \coq's metatheory, as our \lstinline|word n| types are.
}
\lstinline[language=Coq]|Definition vProp  $\Sigma$ : bi := word 64 -b> iPropI  $\Sigma$|.
This is the (\textsc{Coq}) type of assertions in our logic.
Most constructs in Iris's base logic are defined with respect to any BI-algebra (of \textsc{Coq} type \lstinline|bi|), so automatically
carry over to our derived logic.
However, we must still build up from existing Iris primitives to provide new primitives that depend on the address space --- primarily the notion
of virtual points-to.

@@

% \subsection{Base Points-To Assertions}
% \label{sec:pointsto}
% We build up our context-dependent assertions from two basic types of assertions that do not care about address spaces.
To define and use virtual points-to assertions, we require two more basic assertions:

@@

\paragraph{Physical memory  points-to} The soundness proofs for our logic's rules largely center around 
proving that page-table-walk accesses as in Figure \ref{fig:pagetables} succeed, which requires assertions
dealing with physical memory locations.
We have two notions of physical points-to facts. The primitive notion closest to our machine model is captured by an assertion
$ \textsf{pfn} \ \sim \ \textsf{pageoff} \mapsto_{\textsf{a}} \; \{\textsf{q}\} \; \textsf{v} $, where \textsf{pfn} (a $\mathcal{W}_{52}$ \emph{page frame number}) essentially selects a 4KB page of physical memory,
and \textsf{pageoff} (a $\mathcal{W}_{12}$) is an offset within that page.
% could be an 52-bits masked address to level 4 table 
% ($\textsf{w1 } =( \textsf{ l4M52 maddr cr3val}) $),
%and, expectedly \textsf{w2} is an address computed by page-offset computation (e.g. $\textsf{l4off maddr cr3val}$). 
From this we can derive a more concise physical points-to when the split is unimportant:\\
% Giving a raw 64-bits memory pointsto assertion becomes
\centerline{$\textsf{w} \mapsto_{\textsf{p}} \{q\} \textsf{ v} \stackrel{\triangle}{=} (\textsf{drop 12}~w) \ \sim \ (\textsf{bottom 12}~w)\mapsto_{\textsf{a}} \; \{\textsf{q}\} \textsf{ v} $}

 %I put a newline in between the following two Definitions as the second Definition seems not proper without the newline
\begin{figure*}
  \begin{lstlisting}[language=Coq]
 Definition $\vaddr\mapsto_{\textsf{t}}\{\textsf{q}\}\; \vpage$ : vProp $\Sigma$ := 
 $\exists_{\textsf{l4e l3e l2e l1e}} \ldotp$  $\ulcorner$ aligned $\vaddr \urcorner \ast$ L4_L1_PointsTo($\vaddr$ l4e l3e l2e l1e paddr) $\ast$ paddr $\mapsto_{p}\{\mathsf{q}\} \vpage$.
 Definition L4_L1_PointsTo (maddr l4e l3e l2e l1e paddr :word 64) : vProp $\Sigma$ := $\lambda$ cr3val.
  $\ulcorner$ entry_present l4e $\land$ entry_present l3e $\land$ entry_present l2e $\land$ entry_present l1e$\urcorner$ $\ast$
  (l4M52 maddr cr3val) $\sim$ (l4off maddr cr3val) $\mapsto_{a}$ {q1}  l4e  $\ast$
  (l3M52 maddr l4e) $\sim$ (l3off maddr l4e)  $\mapsto_{a}$ {q2}  l3e $\ast$ 
  (l2M52 maddr l3e) $\sim$ (l2off maddr l3e) $\mapsto_{a}$ {q3}  l2e $\ast$
  (l1M52 maddr l2e) $\sim$ (l1off maddr l2e) $\mapsto_{a}$ {q4}  l1e $\ast$ $\ulcorner$ addr_L1(va,l1e) = paddr $\urcorner$.
\end{lstlisting}
\vspace{-1em}
\caption{A Strong Virtual Points-to Relation}
  \label{fig:strongvirtualpointsto}
\vspace{-1em}
\end{figure*}

\subsection{An Overly-Restrictive Definition for Virtual Memory Addressing}
\label{sec:overly-restrictive}
A natural definition for a virtual points-to
asserting that virtual address \textsf{va} points to a value \textsf{v}
would 
% require that in order for a virtual address \textsf{va} to point to a value \textsf{v}, the assertion 
contain
partial ownership of the physical memory involved in the page table walk that would translate \textsf{va} to
its backing physical location --- with locations existentially quantified since a virtual points-to should not assert
\emph{which} locations are accessed in a page table walk, as in Figure \ref{fig:strongvirtualpointsto}.
% Figure \ref{fig:strongvirtualpointsto} gives such a definition.
It asserts the existence of four page table entries, one at each translation level, and via \lstinline|L4_L1_PointsTo|
asserts that the physical page table walk (per Figure \ref{fig:pagetables}) succeeds in reaching the L1 entry,
% asserts from the current \lstinline|cr3| value, taking the offsets from \textsf{va} (as in Figure \ref{fig:pagetables})
% to index each level, finding the existentially quantified entries in each level, 
% ending with the L1 entry, 
% whose physical page base is added to the page offset of \textsf{va} to obtain the physical address
 % --- and there (\lstinline|addr_L1(va,l1e)|) in physical memory, is the value.
which points to the page holding the physical memory backing the virtual address, which contains value \textsf{v}.
Most of the definition lives directly in \textsf{vProp}, using the separation logic structure lifted from Iris's \textsf{iProp}.
\looseness=-1

\lstinline|L4_L1_PointsTo| works by
chaining together the entries for each level, using the sequence of table offsets from the address being translated to index
each table level, and using the physical page address embedded in each entry.\footnote{
  The fractions \lstinline|q1| through \lstinline|q4| represent the fractional ownership of each entry based on how many
  word-aligned addresses might need to share the entry.
}
For example, 
  the first level address translation to get the L4 entry (\lstinline|l4e|) 
  uses the masks \textsf{l4M52} with the current \lstinline|cr3| to get the 
  physical address of the start of the L4 table,
  and \textsf{l4off} with the virtual address being translated to compute the correct offset within that table.
  Then at that physical location is the appropriate entry in the L4 table \textsf{l4e}.
  Subsequent levels of the page table walk work similarly.
The statement of these assertions is simplified by the use of our split physical points-to assertions, since
each level of tables is page-sized.\footnote{We do not address superpages and hugepages in this paper.}
This helper definition is also a more explicit \textsf{vProp}, explicitly binding a value for \lstinline|cr3| and using it
to start the translation process.
\looseness=-1

### 2024-11-12 syntaxsemantics.tex
compressing stuff up through (including) the strong virtual pointsto def

% \subsection{Registers and Memory}
Programs we demonstrate in this paper requires accessing two types of computer resource: registers and memory.
A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$. 
We use these identifiers $\reg$ for register names such as \texttt{rax}, \texttt{r8}, or \texttt{cr3}. Our model includes
all x86-64 integer registers (including stack and instruction pointers), as well as \texttt{cr3} (for page table roots) and \texttt{rflags} (for
flags set by comparison operations and inspected by conditional jumps).
% \begin{figure}[t]
% \newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
% \[
%   \begin{array}{r@{\;}c@{\;}l}
%     \loc & \in & \Loc \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \reg & \in & \regset \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \regval & \in & \regvaltype \\
%     % \val & ::= & \vunit
%   \end{array}
% \quad
%   \begin{array}{r@{\;}c@{\;}l}
%     \instrs & ::= &
%     \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
%     \begin{array}[t]{@{}ll@{}}
%       \iskip
%                    \commentary{no-op}
%       \iseq\instr\instrs
%                    \commentary{sequencing}
%       % \ising\instr
%                    % \commentary{executing}             
%     \end{array}
%     %&
%     %\begin{array}[t]{@{}ll@{}}
%      % \ialloc\lval\allocsize
%       %             \commentary{heap allocation}
%     %\end{array}
%     \end{array}
%     % \\
%     % \ectx & ::= &
%     %   \hole \mid
%     %   \iseq\ectx\instrs 
%     % \\
%   \end{array}
% \]
% \caption{Syntax}
% \Description{Syntax}
% \label{fig:syntax}
% \end{figure}
For clarity and ease of representation, we use machine words, $\loc \in \Loc$, with the subscripts showing the number of bits in a word,
for memory addresses, values, and offsets, rather than distinct location types that wrap machine words.
For example, $\kw{w}_{12}$ is a 12-bit word, which can be obtained for example truncating away 52 bits of a 64-bit word ($\kw{w}_{64}$).

% \subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of word-aligned physical memory addresses 
to 64-bit physical memory values. 
Thus our states $\sigma$ include register maps $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $ and
memory maps $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$.
Of particular note, \texttt{cr3}, the page table register, is included in the machine state.
% As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.

@@

The most important instructions we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right):
\[
\begin{array}{l}
  \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
  \textsf{mov}~[r_m]~r_r \\
  \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
  \textsf{mov}~r_r~[r_m]
  \end{array}
  \]
The semantics of each is realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location
per the description in Section \ref{sec:backgroundonmachinemodel},
then either updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
or loading the value at that physical memory location into the register $r_r$
(\TirNameStyle{WriteToRegFromMem}). 
Additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers) are also treated.
Our formalization also includes basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.

>>>>

% \subsection{Registers and Memory}
% Programs we demonstrate in this paper requires accessing two types of computer resource: registers and memory.
A register identifier, $\reg$, is chosen from a fixed finite set of register identifiers, $\regset$. 
We use these identifiers $\reg$ for register names such as \lstinline|rax|, \lstinline|r8|, or \lstinline|cr3|. Our model includes
all x86-64 integer registers (including stack and instruction pointers), as well as \lstinline|cr3| (for page table roots) and \lstinline|rflags| (for
flags set by comparison operations and inspected by conditional jumps).
% \begin{figure}[t]
% \newcommand{\commentary}[1]{ & \text{\small\it #1} \\}
% \[
%   \begin{array}{r@{\;}c@{\;}l}
%     \loc & \in & \Loc \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \reg & \in & \regset \\
%   \end{array}
%   \begin{array}{r@{\;}c@{\;}l}
%     \regval & \in & \regvaltype \\
%     % \val & ::= & \vunit
%   \end{array}
% \quad
%   \begin{array}{r@{\;}c@{\;}l}
%     \instrs & ::= &
%     \begin{array}[t]{@{}l@{\hspace{10mm}}l@{}}
%     \begin{array}[t]{@{}ll@{}}
%       \iskip
%                    \commentary{no-op}
%       \iseq\instr\instrs
%                    \commentary{sequencing}
%       % \ising\instr
%                    % \commentary{executing}             
%     \end{array}
%     %&
%     %\begin{array}[t]{@{}ll@{}}
%      % \ialloc\lval\allocsize
%       %             \commentary{heap allocation}
%     %\end{array}
%     \end{array}
%     % \\
%     % \ectx & ::= &
%     %   \hole \mid
%     %   \iseq\ectx\instrs 
%     % \\
%   \end{array}
% \]
% \caption{Syntax}
% \Description{Syntax}
% \label{fig:syntax}
% \end{figure}
For clarity and ease of representation, we use machine words, $\loc \in \Loc$, with the subscripts showing the number of bits in a word,
for memory addresses, values, and offsets, rather than distinct location types that wrap machine words.
For example, $\kw{w}_{12}$ is a 12-bit word, which can be obtained for example truncating away 52 bits of a 64-bit word ($\kw{w}_{64}$).

% \subsection{State}
\label{sec:state}
We represent the machine state mainly as a finite map of registers to register values and a map of word-aligned physical memory addresses 
to 64-bit physical memory values. 
Thus our states $\sigma$ include register maps $\sigma.\mathcal{R}: \kw{greg} \rightarrow_{\textrm{fin}} \kw{regval} $ and
memory maps $\sigma.\mathcal{M}: \Locft \rightharpoonup_{\textrm{fin}} (\Loctw \rightharpoonup_{\textrm{fin}} \Locsf )$.
Of particular note, \lstinline|cr3|, the page table register, is included in the machine state.
% As one might have already anticipated from the syntax we introduce, we do not bind any value of an evaluated expression. All the indices and accessed values are treated as globally referenced. In align with this design choice, our expression is a stream of instructions, which is not evaluated to a value to be bound, but changes the machine state through the indices (e.g. $\kw{r}\in\kw{greg}$) -- to the global maps.

@@

The most important instructions we model are memory-accessing variants of the x86-64 \lstinline|mov| instruction, which we format in Intel syntax
(destination on the left, source on the right).
Thus a store to memory is $\textsf{mov}~[r_m]~r_r$, and a load from memory is $\textsf{mov}~r_r~[r_m]$.
% \[
% \begin{array}{l}
%   \hbox{(\TirNameStyle{WriteToMemFromReg})} \qquad
%   \textsf{mov}~[r_m]~r_r \\
%   \hbox{(\TirNameStyle{WriteToRegFromMem})} \qquad 
%   \textsf{mov}~r_r~[r_m]
%   \end{array}
%   \]
Operationally, each first translates the virtual address stored in register $r_m$ to a physical address,
then either updates the memory at that physical location with the contents of $r_r$ (for store) or
or updates $r_r$ with the contents of that memory (for load).
% The semantics of each is realized by first translating the virtual-memory address of a memory location stored in $r_m$ to a physical memory location
% per the description in Section \ref{sec:backgroundonmachinemodel},
% then either updating those memory contents with the contents of register $r_r$ (\TirNameStyle{WriteToMemFromReg}) 
% or loading the value at that physical memory location into the register $r_r$
% (\TirNameStyle{WriteToRegFromMem}). 
Our formalization includes
additional \lstinline|mov| variants (e.g., accessing memory at constant offsets from the base register, or moves between registers),
basic integer and bitwise operations (\lstinline|add|,
\lstinline|and|, bit shifts, etc.) with their effects on \lstinline|rflags|, jumps and some (not all) conditional jumps, \lstinline|call|, \lstinline|ret|, \lstinline|push|,
and \lstinline|pop|.
The semantics model updates to the instruction pointer, though we elide that register (\lstinline|rip|)
from specifications in the paper for clarity.
